The five pillars of computational reproducibility

### 2023-05-19 docs/5pillars.Rmd
data sharing

```{r,dvcs,fig.cap="Figure 3. Distributed version control. Adapted from @Chacon2014-ys.",echo=FALSE,fig.height=6}

@@

It is not sufficient to simply use these tools, they need to be shared as part of the publication
process.
Conda environments are described by a file called `environment.yml` which acts like a recipe for how to
make the environment.
As they are a recipe, these are small files which are easily shared in the project code repository.
Docker images are larger, and cannot be shared in the code repository.
They are commonly deposited to the DockerHub repository, but availability of images is dependent on the
commercial viability of Docker Inc, so it isn't guaranteed to survive the 10-year challenge.
A better option for bioinformatics workflows would be to use the BioContainers registry
@Da_Veiga_Leprevost2017-yw, as it is community driven and supported by consortia such as *Elixir* and
*Global Alliance for Genomics and Health* that will underpin its longevity.
Still, these are not considered permanent archival services, so it is advised to deposit the image
used in anarticle to a suitable persitent long-term archive (discussed below).
The recipe used to build the Docker image, called the Dockerfile, is sufficiently small and can be added
to the project repository.

@@

@@

>>>>

```{r,dvcs,fig.cap="Figure 3. Distributed version control allows. Adapted from @Chacon2014-ys.",echo=FALSE,fig.height=6}

@@

It is not sufficient to simply use these tools, they need to be shared as part of the publication
process.
Conda environments are described by a file called `environment.yml` which acts like a recipe for how to
make the environment.
As they are a recipe, these are small files which are easily shared in the project code repository.
Docker images are larger, and cannot be shared in the code repository.
They are commonly deposited to the DockerHub repository, but availability of images is dependent on the
commercial viability of Docker Inc, so it isn't guaranteed to survive the 10-year challenge.
A better option for bioinformatics workflows would be to use the BioContainers registry
@Da_Veiga_Leprevost2017-yw, as it is community driven and supported by consortia such as *Elixir* and
*Global Alliance for Genomics and Health* that will underpin its longevity.
Still, these are not considered permanent archival services, so it is advised to deposit the image
used in an article to a suitable persitent long-term archive (discussed below).
The recipe used to build the Docker image, called the `Dockerfile`, is sufficiently small and can be
added to the project source code repository.

@@

Without data sharing, computational research is not reproducible nor auditable.
Data sharing is also one of the key features of "open science", which is characterised as science
that is collaborative, transparent, accessible and inclusive [@noauthor_2023-uq].

The Universal Declaration of Human Rights states [@United_Nations1948-ua]:

>Everyone has the right freely to participate in the cultural life of the
>community, to enjoy the arts and to share in scientific advancement and
>its benefits.

In this context, access to research data may be interpreted as a benefit of scientific advancement.
In particular, when large research projects are funded by the public, the public might expect that the
benefts of that study, like the publications and underlying data are made publicly accessible.

Data sharing has dual purposes, firstly to facilitate reproduction and auditing, and secondly to
enable reuse in other contexts [@Wilkinson2016-qq].
Data reuse increases efficiency, as it prevents redundant research and facilitates new research ideas
that were impossible without it.
In terms of research rigour, inspecting raw data can uncover research integrity problems [@Miyakawa2020-yj],
in addition to validating scientific findings by independent computational reproduction [@Ioannidis2009-hv].
In light of this, a research article without supporting data and code is much like a press release or
advertisement where the claims made cannot be verified as true or not.

Although it is common to see journal articles with "Data available upon reasonable request", this is less
than ideal.
A systematic study of such data availability statements found that the data was successfully provided
in just 6.8% of requests [@Gabelica2022-ci].
In the rare cases that data is shared in accordance with the data availability statement, it poses
a burden in terms of labour to lodge and respond to such requests, and if the data is large, may incur
an additional cost for hosting/transferring these large files.

The field of genomics has has a long history of data sharing @Brazma2001-yk, which is supported by
policy leadership by funding bodies
[@National_Health_and_Medical_Research_Council2019-yx;
@Office_of_The_Director_National_Institutes_of_Health2023-ss] and data sharing policies of journals
(eg [@Packer2018-ru; @Hanson2011-ub]), although there are some challenges.

Some researchers would rather keep data for their private use.
Some may keep data private due to concerns over participant privacy.
This is a well-founded concern for identifiable data, however once identifying features are removed,
it is nearly impossible to reidentify genomic data @Venkatesaramani2021-tn unless the attackers have
preexisting genomic data describing an individual or ancestry [@Schadt2012-md; @Gymrek2013-fz].
Therefore sharing deidentified genomic data and metadata is likely to carry little risk to participant
privacy.
Still, much human genomic data and metadata resides in restricted databases due to the scope of
data sharing that was agreed to in the consent process.
For example the consent form may only allow sharing of data to researchers investigating certain
diseases, which means a vetting process must be undertaken before sharing data with external
investigators.
In the interests of reuse and reproducibility, future studies could consider a consent form that allows
more broad sharing of deidentified data.

In order to maximise the value of shared data for reuse and reproducibility, it needs to be Findable,
Accessible, Interoperable and Reusable (FAIR) for people and for computers @Wilkinson2016-qq.
High throughput biology has already built a number of data repositories for specific data types
conforming to FAIR principles (eg: GEO, SRA, PRIDE [@Edgar2002-hs; @Kodama2012-bn; @Deutsch2017-hc]).
However other domains lack well known specialised organised data repositories, such as evolutionary
biology, ecology, cell biology, social sciences and computing.
The re3data.org registry of data repositories may be useful to find a repository that accepts data
from these fields [@Pampel2013-ep].
Moreover general recommendations have been proposed to enhance data FAIRness in ecology, but which
can also be applied to other disciplines [@Roche2015-aq].

Key recommendations include:

1. Depositing data to a specialised repository if possible, otherwise a general purpose repository.

2. If possible, avoid large supplementary data files accompanying journal articles.

2. Provide detailed metadata.

3. Preferably archive and share raw data, and use existing standards for the discipline.

Once the quality of the data set has been established researchers may consider depositing it
to a data repository early, before publicisation, as it has some advantages.
Most data repositories have an optional 12-month embargo period so researchers can share publicly
at a later date once they're happy with the overall project.
The data repository acts as an additional backup to the host institutions' own in case of data loss
or calamity.

By assuming the data and metadata can be sourced from a public location, the workflow developers ensure
that the code they share won't suffer from "file not found errors" during reproduction.
This ensures code and data are linked, which has been a  [@Peng2011-ia]

@@

* Dryad

* GigaDB
